Automatic Parallelization and Transparent Fault Tolerance\\ (New project paper)

### 2016-04-05 pubs/tfp16/paper.tex
more small changes

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
\usepackage{comment}
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Automatic Parallelization and Transparent Fault Tolerance\\
       (New project paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\subsection{The trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new, as evidenced
by the Workshop on Functional High Performance Computing, held in conjuction
with the International Conference on Functional Programming, being now in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of these ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

@@

\subsubsection{Proper tail calls.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009}.  To achieve efficient parallel execution they
identified the need for the ability to examine and unwind the stack.  This
requires that use of the C stack be highly constrained if we are to 
avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
non-portable code to directly manipulate the C stack, complicating an
already non-trivial implementation effort.  We therefore must limit the
use of the C stack to calls to the runtime, in other words, when executing
code generated from the user program (and not in a runtime call), there
must be no growing of (pushing return addresses or data onto) the C stack.

@@

There are several hundred small STG and mini-Haskell programs in the test suite.
It is partitioned into a number of directories which control what
testing is done for a specific program. Given our goal of exploring various
levels of strictness, most tests are run with all evaluation strategies (three
degrees of function strictness by two degrees of constructor strictness).
One interesting class of test programs are those which require lazy
semantics. For these programs we test not only that the test leads to the expected
result with non-strict semantics, but also that the program returns a
\emph{BLACKHOLE} in the case of strict semantics. Students who were Haskell/STG
beginners have contributed substantially to the test suite. We have also found
that the test suite has be of use as a learning tool for these students to
discover what level of non-strictness was required for their programs,
a benefit that we did not initially envision. Over time we have developed
extensive STG and mini-Haskell preludes of common programming patterns to
aid in the writing of test programs.

@@

debug level (runtime/compile time)

@@

Parallel runtime is being implemented following the ideas
of~\cite{SPJs-many-papers}, thread pool, stack and heap for each thread, blah,
blah.

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
%
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
\usepackage{comment} % comment environment
\usepackage{hyperref} 
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Automatic Parallelization and Transparent Fault Tolerance\\
       (New project paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\subsection{The trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new, as evidenced
by the Workshop on Functional High Performance Computing, held in conjunction
with the International Conference on Functional Programming, being now in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of these ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

@@

\subsubsection{Proper tail calls.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009,Marlow:2011}.  To achieve efficient parallel 
execution they identified the need for the ability to examine and unwind the 
stack.  This requires that use of the C stack be highly constrained if we are 
to avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
non-portable code to directly manipulate the C stack, complicating an
already non-trivial implementation effort.  We therefore must limit the
use of the C stack to calls to the runtime, in other words, when executing
code generated from the user program (and not in a runtime call), there
must be no growing of (pushing return addresses or data onto) the C stack.

@@

There are several hundred small STG and mini-Haskell programs in the test suite 
(Most of which were written by students).
It is partitioned into a number of directories which control what
testing is done for a specific program. Given our goal of exploring various
levels of strictness, most tests are run with all evaluation strategies (three
degrees of function strictness by two degrees of constructor strictness).
One interesting class of test programs are those which require lazy
semantics. For these programs we test not only that the test leads to the expected
result with non-strict semantics, but also that the program returns a
\emph{BLACKHOLE} in the case of strict semantics. Students who were Haskell/STG
beginners have contributed substantially to the test suite. We have also found
that the test suite has be of use as a learning tool for these students to
discover what level of non-strictness was required for their programs,
a benefit that we did not initially envision. Over time we have developed
extensive STG and mini-Haskell preludes of common programming patterns to
aid in the writing of test programs.

@@

We have also implemented a simple logging facility that can be controlled at 
both compile and runtime which helps in debugging why a test is failing.

@@

Parallel runtime is being implemented following the ideas
of~\cite{Harris:2005,Marlow:2008,Marlow:2009}, thread pool, stack and heap 
for each thread, blah, blah.
